---
title: "Homerwork 2"
author: "Eugene Nesterenko"
date: 2023-05-21
format: 
  docx: default
  html:
    toc: true
    toc_float: true
    code-fold: true
editor: visual
---

```{r}
#| label: load-libraries
#| echo: false # This option disables the printing of code (only output is displayed).
#| message: false
#| warning: false

library(tidyverse)
library(wbstats)
library(skimr)
library(countrycode)
library(here)
```

# Data Visualisation - Exploration

Now that you've demonstrated your software is setup, and you have the basics of data manipulation, the goal of this assignment is to practice transforming, visualising, and exploring data.

# Mass shootings in the US

In July 2012, in the aftermath of a mass shooting in a movie theater in Aurora, Colorado, [Mother Jones](https://www.motherjones.com/politics/2012/07/mass-shootings-map/) published a report on mass shootings in the United States since 1982. Importantly, they provided the underlying data set as [an open-source database](https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/) for anyone interested in studying and understanding this criminal behavior.

## Obtain the data

```{r}
#| echo: false
#| message: false
#| warning: false


mass_shootings <- read_csv(here::here("data", "mass_shootings.csv"))

glimpse(mass_shootings)
```

| column(variable)     | description                                                                 |
|--------------------------|----------------------------------------------|
| case                 | short name of incident                                                      |
| year, month, day     | year, month, day in which the shooting occurred                             |
| location             | city and state where the shooting occcurred                                 |
| summary              | brief description of the incident                                           |
| fatalities           | Number of fatalities in the incident, excluding the shooter                 |
| injured              | Number of injured, non-fatal victims in the incident, excluding the shooter |
| total_victims        | number of total victims in the incident, excluding the shooter              |
| location_type        | generic location in which the shooting took place                           |
| male                 | logical value, indicating whether the shooter was male                      |
| age_of_shooter       | age of the shooter when the incident occured                                |
| race                 | race of the shooter                                                         |
| prior_mental_illness | did the shooter show evidence of mental illness prior to the incident?      |

## Explore the data

### Specific questions

-   Generate a data frame that summarizes the number of mass shootings per year.

```{r}
# Count the number of mass shootings per year
mass_shootings %>% 
  group_by(year) %>% 
  summarise(number = n())
```

-   Generate a bar chart that identifies the number of mass shooters associated with each race category. The bars should be sorted from highest to lowest and each bar should show its number.

```{r}

# Counting number of shootings per race,excluding cases with unidentified race (NA)
mass_shootings %>%
  filter(!is.na(race)) %>% 
  group_by(race) %>% 
  summarise(number = n()) %>% 
  
  # Rearranging in descending order for total number
  mutate(race = fct_reorder(race, number)) %>%
  
  # Drawing a horizontal bar chart with labels
  ggplot(aes(x = number, y = race)) +
  geom_col() +
  geom_text(
    aes(label = number, x = number - .25),
    colour = "white",
    size = 5,
    hjust = 1) +
  
  # Adding theme and axis titles
  theme_minimal(base_size = 12)+
  labs(
        title = "",
        subtitle = "",
        x = "Number of shootings",
        y = "Race"
      ) +
      NULL
```

-   Generate a boxplot visualizing the number of total victims, by type of location.

```{r}

# Grouping shootings by location type, excluding those with unidentified location (NA)
mass_shootings %>%
  filter(!is.na(location_type)) %>% 
  group_by(location_type) %>% 
  
  # Drawing a boxplot
  ggplot(aes(y = total_victims, x = location_type)) +
  geom_boxplot() +
  
  # Drawing y-axis (number of shootings) in a log scale to interpret graph more easily
  scale_y_log10() +
  
  # Adding theme and and axis titles
  theme_classic(base_size = 12) +
  labs(
        title = "",
        subtitle = "",
        x = "Type of location",
        y = "Number of shootings"
      ) +
      NULL
```

-   Redraw the same plot, but remove the Las Vegas Strip massacre from the dataset.

```{r}

# Grouping shootings by location type, excluding those with unidentified location (NA)
mass_shootings %>%
  filter(!is.na(location_type)) %>%
  
  # Excluding the Las Vegas Strip massacre
  filter(case != "Las Vegas Strip massacre") %>% 
  group_by(location_type) %>% 
  
  # Drawing a boxplot
  ggplot(aes(y = total_victims, x = location_type)) +
  geom_boxplot() +
  
  # Drawing y-axis (number of shootings) in a log scale to interpret graph more easily
  scale_y_log10() +
  
  # Adding theme and and axis titles
  theme_classic(base_size = 12) +
  labs(
        title = "",
        subtitle = "",
        x = "Type of location",
        y = "Number of shootings"
      ) +
      NULL
```

### More open-ended questions

Address the following questions. Generate appropriate figures/tables to support your conclusions.

-   How many white males with prior signs of mental illness initiated a mass shooting after 2000?

```{r}

# Selecting only cases after 2000, where the shooter was white male with prior signs of mental illness
mass_shootings %>%
  filter(race == "White") %>%
  filter(male == TRUE) %>% 
  filter(prior_mental_illness == "Yes") %>% 
  filter(year > 2000) %>% 
  
  # Counting total number of cases after applyied filters
  summarise(count = n())
```

-   Which month of the year has the most mass shootings? Generate a bar chart sorted in chronological (natural) order (Jan-Feb-Mar- etc) to provide evidence of your answer.

```{r}

# We have not found a function in lubridate library, so we will create month index manually
mass_shootings_month <- mass_shootings %>% 
  mutate(month_index = case_when(
    month == "Jan" ~1,
    month == "Feb" ~2,
    month == "Mar" ~3,
    month == "Apr" ~4,
    month == "May" ~5,
    month == "Jun" ~6,
    month == "Jul" ~7,
    month == "Aug" ~8,
    month == "Sep" ~9,
    month == "Oct" ~10,
    month == "Nov" ~11,
    month == "Dec" ~12,
    TRUE ~0
   )
  )

# Count cases of mass shooting by month and save as new data frame
mass_shootings_month_chronological_summary <- mass_shootings_month %>% 
  group_by(month, month_index) %>%
  summarise(count = n()) %>% 
  
  # We arrange in chronological order using month index (Jan = 1, Dec = 12)
  arrange(month_index)


# Attempt to arrange bar chart in chronologocal order usung month index (unsuccessful)
mass_shootings_month_chronological_summary %>% 
  slice_max(order_by = month_index, n=12) %>% 

  
# Drawing a vertical bar chart with labels
  ggplot(aes(x = month, y = count)) +
  geom_col() +
  
  # Adding theme and axis titles
  theme_minimal(base_size = 12) +
  labs(
        title = "",
        subtitle = "",
        x = "Number of shootings",
        y = "Month"
      ) +
      NULL
```
